---
title: "tidyverse_slides"
author: "Edzer Pebesma"
date: "6/3/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidyverse

What is tidyverse? It's a word play, a contraction of [tidy data](https://www.jstatsoft.org/article/view/v059i10) and the
_Hadleyverse_, a universe of packages created by Hadley Wickham. (The word _tidyverse_ was chosen by Hadley)

Tidy also refers to the [tidy tools manifesto](https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html), a vignette in the _tidyverse_ package. 
Quoting: "There are four basic principles to a tidy API:

1.  Reuse existing data structures.
1. Compose simple functions with the pipe.
1.  Embrace functional programming.
1. Design for humans."

Tidyverse is also a meta-package, a package that loads a lot of other packages:
```{r}
library(tidyverse)
```
If you worry about dependencies, you should never write a package that depends on a metapackage:
```{r}
devtools::session_info()
```

## `tibble`

A `tibble` is an evolution of the `data.frame`:
```{r}
(tb = tibble(a = 3))
```
It "is" a `data.frame`,
```{r}
inherits(tb, "data.frame")
```
but behaves differently
```{r}
tb[,"a"] # defaults to drop = FALSE
as.data.frame(tb)[, "a"] # drop = TRUE
```
Most people will agree that tibbles and `data.frame`s are different, many will say the difference is not very important.
Some people (like me) will say it's useful to understand the difference.

## `readr`

Tidyverse tries to consistently use `_` in function names.

Type `read_` and pres `TAB` twice, to see the list of `readr` functions starting with `read_`:
```
read_csv                read_delim_chunked      read_lines_chunked      read_table
read_csv_chunked        read_file               read_lines_raw          read_table2
read_csv2               read_file_raw           read_lines_raw_chunked  read_tsv
read_csv2_chunked       read_fwf                read_log                read_tsv_chunked
read_delim              read_lines              read_rds    
```

These functions may all have their reasons to be prefered over there `read.xxx` counterpart in base R. One thing to be aware of is that they all read character data as `character` vectors, where (by default) `read.csv` and `read.table` etc. read character data into `factor` variables, unless you set (globally -- never do this! --, or at the start of your script -- better! --)
```{r}
options(stringsAsFactors = FALSE)
```
Also, all these functions will return `tibble`s, rather than `data.frame`s.

## `dplyr`

The work horse of the tidyverse. Key commands:

1. `filter` selects records
2. `select` selects variables
3. `mutate` adds/changes variables
4. `left_join`, `inner_join`, `full_join` etc
5. `group_by` creates a _grouped_ tibble
6. `summarise` summarises variables, by group if grouped

```{r}
library(units)
(lkm = set_units(1, miles/gallon) %>% set_units(l/km))
mtcars %>%
  filter(cyl <= 8)	%>%
	mutate(usage = mpg * lkm) %>%
  group_by(cyl) %>%
	summarize(mean(usage), mean(mpg))
```

## `dbplyr`: database connectivity

The number one killer reason to use tidyverse, and maybe even to use R rather than Python. 

1. develop your workflow on a small table, using `dplyr`
2. open a proxy object to a database in `dbplyr`
3. run the same workflow/script against that object
4. voila. Everything is done _inside_ the database.

DBI/Spark/Cloudera/BigQuery/...

## `tidyr`

Reshapes tables: we saw the need for `gather` (opposit of `spread`) when plotting multiple attributes with `geom_sf`.
These commands will be renamed into `pivot_long` and `pivot_wide` in the future.

Other activities: `nest`, and `unnest` to nest tibbles (have a tibble as a record in a list-column). Example: tidy storm [trajectories](https://www.r-spatial.org/r/2017/08/28/nest.html)

## `stringr`

Helps with handling string (`character`) data. (Another package, `stringi`, does that too.)

## `forcats`

Helps with the handling of `factors`.

## `purrr`

functional programming: alternatives to `lapply`, `sapply`, `vapply`, `apply` and so on. We will talk about functional programming.

## Other packages

Many packages work along / well with `tidyverse`. One of them is `sf`:
```{r}
library(sf)
methods(class = 'sf')
```